Water desalination using polyelectrolyte hydrogel. Gibbs ensemble modeling

### 2022-07-08 esi.tex
Edited donnan section in esi

\subsection{Donnan equilibrium in finite volume reservoirs}
\todo{mb we remove this section}
Before we can explore any system properties and collect data the system has to be equilibrated. To achieve this we perform equilibration routine consist of sequence of Molecular Dynamics and Monte Carlo steps.
%Here we describe our technique
The result of equilibration is stationary of any observable such as pressure, end-to-end distance for polymer chains and ion distribution between reservoirs.

The success and the cost in terms of computational time for the equilibration process heavily depend on initial guess. In our case we initialize the system with ion distribution corresponds to Donnan equilibrium.

In closed system number of ions in the system is constant, so Donnan equilibrium can be formulated as follows
\begin{equation}
    \frac{\left(N_{pairs} - N_{Cl^{-}}^{(gel)}\right)^2}{V_{salt}^2} = \frac{N_{Cl^{-}}^{gel} (N^{{(gel)}}_{A^{-}} + N_{Cl^{-}}^{(gel)})}{V_{gel}^2}
\end{equation}
where $N^{{(gel)}}_{A^{-}}$ is number of fixed anions in the box with the gel \ie acetic groups in the polymer gel.
Solving this equation with respect to $N^{{(gel)}}_{A^{-}}$ gives us the distribution of the ions between two boxes.

\begin{equation}
N_{Cl^{-}}^{(gel)} = \frac{\frac{\NA \Vout^{2}}{2} + N_{pairs} \Vgel^{2} - \frac{\Vout \sqrt{N^{{(gel)}}_{A^{-}}^{2} \Vout^{2} + 4 N^{{(gel)}}_{A^{-}} N_{pairs} \Vgel^{2} + 4 N_{pairs}^{2} \Vgel^{2}}}{2}}{\Vgel^{2} - \Vout^{2}}
\end{equation}

\todoi{
This paragraph moved from overleaf

Compression of the gel in infinite reservoir does not  change its salinity, on the contrary, upon compression of the gel in closed reservoir salinity of separated liquid is decreasing (supernatant salinity), using the Donnan law formulated for finite reservoir salinity can be expressed as
\begin{equation*}
    c_s^\prime = \frac{\sqrt{4\alpha^{2} c_{p}^{2} \left(1-\gamma\right)^{2} + \alpha c_{p} c_{s} \gamma^{2} + c_{s}^{2} \gamma^{2}}-\frac{\alpha c_{p} \left(1-\gamma\right)}{2} - c_{s} \left(1-\gamma\right)}{\gamma^{2} - \left(1-\gamma\right)^{2}}
\end{equation*}
where $\gamma = V_{gel}/V_0$ is relative to initial volume compression.
}
\begin{eqnarray}
    %N_{Cl^{-}}^{(gel)} = \frac{\frac{N^{{(gel)}}_{A^{-}} V_{salt}^{2}}{2} + N_{pairs} V_{gel}^{2} - \frac{V_{salt} \sqrt{N^{{(gel)}}_{A^{-}}^{2} V_{salt}^{2} + 4 N^{{(gel)}}_{A^{-}} N_{pairs} V_{gel}^{2} + 4 N_{pairs}^{2} V_{gel}^{2}}}{2}}{V_{gel}^{2} - V_{salt}^{2}}
    \\
    N_{Na^{+}}^{(gel)} = N^{{(gel)}}_{A^{-}} + N_{Cl^{-}}^{(gel)}
    \\
    N_{Cl^{-}}^{(salt)} = N_{Na^{+}}^{(gel)} = N_{pairs} - N_{Cl^{-}}^{(gel)}
\end{eqnarray}
\subsection{Sampling routine}
Data gained from Molecular Dynamics (MD) and Monte Carlo processes (MC) are highly autocorrelated, thus we can not imply statistic developed for uncorrelated data. 
We have written a small in-house python package realizing the routine estimating mean and error for the observables provided by simulation, \ie pressure, number of particles, end-to-end distance of a chain for a given confidence level or effective sample size. 
Within this routine we provided the way to limit sampling procedure with some timeout value, in a way similar to described in \href{https://www.physik.uni-leipzig.de/~janke/Paper/nic10_423_2002.pdf}{Janke2002}

>>>>

\subsection{Donnan equilibrium in finite volume reservoirs}
Before we can explore any system properties and collect data the system has to be equilibrated. To achieve this we perform equilibration routine which consist of sequence of Molecular Dynamics and Monte Carlo steps.
The result of equilibration is stationary of any observable such as pressure, end-to-end distance for polymer chains and ion distribution between reservoirs. 
The success and the cost in terms of computational time for the equilibration process heavily depend on initial guess. 
In our case we initialize the system with ion distribution corresponds to Donnan equilibrium.

In closed system number of ions in the system is constant, so Donnan equilibrium can be formulated as follows
\begin{equation}
    \frac{\Ncl\out\Nna\out}{\Vout^2} = \frac{\Ncl\gel\Nna\gel}{\Vgel^2}
\end{equation}
or
\begin{equation}
    \frac{\left(\Ncl\tot - \Ncl\gel\right)^2}{\Vout^2} = \frac{\Ncl\gel (\NA + \Ncl\gel)}{\Vgel^2}
\end{equation}
where $\NA$ is number of fixed anions in the box with the gel \ie acetic groups in the polymer gel.
Solving this equation with respect to $\Ncl\gel$ gives us the distribution of the ions between two boxes.
\begin{equation}
\frac{- N - \frac{1}{2}\NA \gamma^{2} + \gamma \NA \sqrt{\left(\frac{N}{\NA}\right)^{2} + \frac{N}{\NA} +  \left(\frac{\gamma}{2}\right)^{2}}}{\gamma^{2} - 1 }
\end{equation}
where $\gamma = \Vout/\Vgel$ is a relative to initial volume compression.

\subsection{Sampling routine}
Data gained from Molecular Dynamics (MD) and Monte Carlo processes (MC) are highly autocorrelated, thus we can not imply statistic developed for uncorrelated data. 
We have written a small in-house python package realizing the routine estimating mean and error for the observables provided by simulation, \ie pressure, number of particles, end-to-end distance of a chain for a given confidence level or effective sample size. 
Within this routine we provided the way to limit sampling procedure with some timeout value, in a way similar to described in \href{https://www.physik.uni-leipzig.de/~janke/Paper/nic10_423_2002.pdf}{Janke2002}

### 2022-07-08 gibbs.tex
Edited donnan section in esi

\paragraph{Physics behind the desalination process.}
%The maximum entropy principle requires the density of charges to be more-less the same throughout the system medium. 
Since the polyelectrolyte gel has its own charges and its own neutralizing counterions, the density of mobile ions (which are able to freely enter and leave the gel) in gel appears to be smaller than their density outside the gel.
Therefore the internal solution of the gel has a lower density of mobile ions than that in the solution outside. 
In that sense, the gel acts as an osmotic membrane separating solutions \cite{Rud2020}. 
The driving force of the separation is the Donnan potential originated from the charges of the hydrogel network. 
The density of mobile ions difference between ions concentrations in the internal and external solutions
is defined by Donnan law \cite{Rud2018}
\begin{equation*}
\frac{\ccl\gel}{\cs}=\frac{\cs}{\cna\gel}=\sqrt{1+\left(\frac{\alpha}{2\cs\Vgel}\right)^{2}}\pm\frac{\alpha}{2\cs\Vgel}
\end{equation*}
where $\ccl\gel$ and $\cna\gel$ are the internal concentration of monovalent anions and cations, $\cs$ is the salinity of external solution $\Vgel$ is the gel molar volume (inverse density of gel segments) and $\alpha$ is ionization degree. 
The $\pm$ sign in the formula accounts for the change between the cases of gel made of polyanion or polycation.

@@

In order to get the partial pressure of the gel, we substitute from $P$ the osmotic pressure of ions in the reservoir: $\Pgel=P - \Pout$.
$\Pout$, in turn, we calculate by running a separate simulation of a reservoir, containing ionic gas in equilibrium with the bath of the same salinity $\cs$. 
The gel partial pressure, $\Pgel$, is the pressure that needs to be applied to the gel via a solvent permeable filter to compress the gel to a specific molar volume, $\Vgel$. 
By gel molar volume $\Vgel$ we assume the volume of the gel calculated per on mole of gel monomer units.

@@

\multicolumn{1}{l|}{$c_{s}^{0}$, mM} & $c_{s}^{5}$, mM & \multicolumn{1}{l|}{$n_{\cl}^{0}$, mM} & $n_{\cl}^{5}$, mM & $\Delta v$, l & $|W$\textbar , J/L & \multicolumn{2}{c}{$W^{id}$, J/L}\tabularnewline

>>>>

\paragraph{Physics behind the desalination process.}
%The maximum entropy principle requires the density of charges to be more-less the same throughout the system medium. 
Since the polyelectrolyte gel has its own charges and its own neutralizing counterions, the density of mobile ions (which are able to freely enter and leave the gel) in gel appears to be smaller than their density outside the gel.
Therefore the internal solution of the gel has a lower density of mobile ions than that in the solution outside. 
In that sense, the gel acts as an osmotic membrane separating solutions \cite{Rud2020}. 
The driving force of the separation is the Donnan potential originated from the charges of the hydrogel network. 
The density of mobile ions difference between ions concentrations in the internal and external solutions
is defined by Donnan law \cite{Rud2018}
\begin{equation*}
\frac{\ccl\gel}{\cs}=\frac{\cs}{\cna\gel}=\sqrt{1+\left(\frac{\alpha}{2\cs\Vgel}\right)^{2}}\pm\frac{\alpha}{2\cs\Vgel}
\end{equation*}
where $\ccl\gel$ and $\cna\gel$ are the internal concentration of monovalent anions and cations, $\cs$ is the salinity of external solution,  $\Vgel$ is the gel molar volume (inverse density of gel segments) and $\alpha$ is ionization degree (in the case of our study $\alpha=1$). 
The $\pm$ sign in the formula accounts for the change between the cases of gel made of polyanion or polycation.

@@

In order to get the partial pressure of the gel, we substitute from $P$ the osmotic pressure of ions in the reservoir: $\Pgel=P - \Pout$.
$\Pout$, in turn, we calculate by running a separate simulation of a reservoir, containing ionic gas in equilibrium with the bath of the same salinity $\cs$. 
The gel partial pressure, $\Pgel$, is the pressure that needs to be applied to the gel via a solvent permeable filter to compress the gel to a specific molar volume, $\Vgel$. 
By gel molar volume $\Vgel$ we assume the volume of the gel calculated per on mol of gel monomer units.

@@

\multicolumn{1}{l|}{$c_{s}^{0}$, mM} & $c_{s}^{5}$, mM & \multicolumn{1}{l|}{$n_{\cl}^{0}$, mM} & $n_{\cl}^{5}$, mM & $\Delta \Vgel$, l & $|W$\textbar , J/l & \multicolumn{2}{c}{$W^{id}$, J/l}\tabularnewline
